---
title: "Wind Effects on Monarch Abundance: Exploratory GAM Analysis"
author: "Kyle Nessen"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 8
    self-contained: true
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 8)

# Load required packages
library(tidyverse)
library(mgcv)      # GAM models
library(gratia)    # GAM diagnostics and plotting
library(lubridate) # Date handling
library(knitr)
library(patchwork) # Plot combinations
library(performance) # Model diagnostics
library(see)       # Enhanced performance plots  
# library(data.table) # For efficient joins

# Set theme for all plots
theme_set(theme_minimal(base_size = 14))
```

# Data Loading and Preparation

```{r load-data}
# Load data files
butterfly_data <- read_csv("data/butterfly_abundance_index.csv")
wind_data <- read_csv("data/wind_all.csv")
temperature_data <- read_csv("data/temperature_data_2023.csv")
deployments <- read_csv("data/deployments.csv")

# Inspect data structure
glimpse(butterfly_data)
glimpse(wind_data)
glimpse(temperature_data)
```

```{r data-preparation}
# Extract datetime from butterfly image filenames
butterfly_data <- butterfly_data %>%
  mutate(
    # Extract timestamp from filename (format: SC1_20231117114001.JPG)
    timestamp_str = str_extract(image_filename, "\\d{14}"),
    datetime = ymd_hms(timestamp_str, truncated = 3),
    date = as_date(datetime),
    hour = hour(datetime)
  )

# Convert wind data datetime
wind_data <- wind_data %>%
  mutate(
    datetime = ymd_hms(time),
    date = as_date(datetime),
    hour = hour(datetime),
    minute = minute(datetime)
  )

# Temperature data preparation
temperature_data <- temperature_data %>%
  mutate(
    datetime = ymd_hms(timestamp),
    date = as_date(datetime),
    hour = hour(datetime)
  )

# Filter for relevant deployments (those with monarchs present AND complete labeling)
relevant_deployments <- deployments %>%
  filter(monarchs_present == TRUE, 
         label_status == "Complete") %>%
  select(deployment_id, view_id, wind_meter_name, Deployed_time, Recovered_time)

# Filter butterfly data to relevant deployments only
butterfly_data <- butterfly_data %>%
  inner_join(relevant_deployments, by = "deployment_id")

print(paste("Number of butterfly observations:", nrow(butterfly_data)))
print(paste("Number of unique deployments:", n_distinct(butterfly_data$deployment_id)))
print(paste("Date range:", min(butterfly_data$date), "to", max(butterfly_data$date)))
```

```{r merge-data}
# Simplified data merging approach
print("Creating merged dataset with wind data...")

# Since butterfly_data was already filtered to complete deployments with wind meters,
# we need to be more efficient about the wind data join to avoid memory issues

# First, let's filter wind data to only relevant time periods and meters
relevant_wind_meters <- unique(butterfly_data$wind_meter_name)
wind_filtered <- wind_data %>%
  filter(wind_meter_name %in% relevant_wind_meters,
         datetime >= min(butterfly_data$datetime) - minutes(60),
         datetime <= max(butterfly_data$datetime) + minutes(60))

print(paste("Filtered wind data to", nrow(wind_filtered), "observations"))

# Now calculate wind summaries more efficiently
wind_summaries <- butterfly_data %>%
  rowwise() %>%
  do({
    current_obs <- .
    current_wind <- wind_filtered %>%
      filter(wind_meter_name == current_obs$wind_meter_name,
             datetime >= current_obs$datetime - minutes(30),
             datetime <= current_obs$datetime)
    
    if(nrow(current_wind) >= 5) {  # At least 5 observations
      tibble(
        deployment_id = current_obs$deployment_id,
        datetime = current_obs$datetime,
        image_filename = current_obs$image_filename,
        mean_wind_speed = mean(current_wind$speed, na.rm = TRUE),
        max_wind_speed = max(current_wind$speed, na.rm = TRUE),
        max_gust_speed = max(current_wind$gust, na.rm = TRUE),
        wind_variance = var(current_wind$speed, na.rm = TRUE),
        minutes_above_2ms = sum(current_wind$speed > 2, na.rm = TRUE),
        gust_minutes_above_2ms = sum(current_wind$gust > 2, na.rm = TRUE),
        n_wind_obs = nrow(current_wind)
      )
    } else {
      tibble(
        deployment_id = current_obs$deployment_id,
        datetime = current_obs$datetime,
        image_filename = current_obs$image_filename,
        mean_wind_speed = NA_real_,
        max_wind_speed = NA_real_,
        max_gust_speed = NA_real_,
        wind_variance = NA_real_,
        minutes_above_2ms = NA_real_,
        gust_minutes_above_2ms = NA_real_,
        n_wind_obs = 0
      )
    }
  }) %>%
  ungroup()

print(paste("Wind summaries calculated for", nrow(wind_summaries), "observations"))

# Join wind summaries back to butterfly data
merged_data <- butterfly_data %>%
  left_join(wind_summaries, by = c("deployment_id", "datetime", "image_filename"))

# Add temperature data
merged_data <- merged_data %>%
  left_join(
    temperature_data %>% 
      select(deployment_id, timestamp, temperature) %>%
      mutate(datetime = ymd_hms(timestamp)) %>%
      select(-timestamp),
    by = c("deployment_id", "datetime")
  )

# Add lagged abundance (previous observation)
merged_data <- merged_data %>%
  arrange(deployment_id, datetime) %>%
  group_by(deployment_id) %>%
  mutate(
    abundance_lag1 = lag(total_butterflies, 1),
    time_since_last = as.numeric(datetime - lag(datetime), units = "hours")
  ) %>%
  ungroup()

# Calculate sunlight exposure proportion
merged_data <- merged_data %>%
  mutate(
    sunlight_prop = ifelse(total_butterflies > 0, 
                          butterflies_direct_sun / total_butterflies, 
                          0)
  )

# Create final analysis dataset
analysis_data <- merged_data %>%
  filter(
    !is.na(mean_wind_speed), 
    !is.na(abundance_lag1),
    !is.na(temperature),
    n_wind_obs >= 5,  # At least 5 wind observations in 30-minute window
    time_since_last <= 12  # Remove observations with gaps > 12 hours
  ) %>%
  mutate(
    # Create time-based variables
    hour_of_day = hour(datetime),
    day_of_year = yday(datetime),
    # Log-transform abundance for better model behavior
    log_abundance = log(total_butterflies + 1),
    log_abundance_lag1 = log(abundance_lag1 + 1),
    # Create factor for view_id
    view_id = factor(view_id)
  )

print(paste("Final analysis dataset:", nrow(analysis_data), "observations"))
print(paste("Deployments in final dataset:", n_distinct(analysis_data$deployment_id)))
print(paste("View IDs in final dataset:", n_distinct(analysis_data$view_id)))
print(paste("Date range:", min(analysis_data$date), "to", max(analysis_data$date)))
```

# Exploratory Data Analysis

```{r eda-summary}
# Summary statistics
analysis_data %>%
  select(total_butterflies, mean_wind_speed, max_wind_speed, wind_variance, 
         temperature, sunlight_prop) %>%
  summary() %>%
  kable(caption = "Summary Statistics of Key Variables")
```

```{r eda-plots}
# Distribution plots
p1 <- ggplot(analysis_data, aes(x = total_butterflies)) +
  geom_histogram(bins = 50, alpha = 0.7) +
  scale_x_log10() +
  labs(title = "Distribution of Butterfly Abundance", 
       x = "Total Butterflies (log scale)", y = "Count")

p2 <- ggplot(analysis_data, aes(x = mean_wind_speed)) +
  geom_histogram(bins = 30, alpha = 0.7) +
  labs(title = "Distribution of Mean Wind Speed", 
       x = "Mean Wind Speed (m/s)", y = "Count") +
  geom_vline(xintercept = 2, color = "red", linetype = "dashed", alpha = 0.7)

p3 <- ggplot(analysis_data, aes(x = temperature)) +
  geom_histogram(bins = 30, alpha = 0.7) +
  labs(title = "Distribution of Temperature", 
       x = "Temperature (°C)", y = "Count")

p4 <- ggplot(analysis_data, aes(x = sunlight_prop)) +
  geom_histogram(bins = 30, alpha = 0.7) +
  labs(title = "Distribution of Sunlight Exposure Proportion", 
       x = "Proportion in Direct Sunlight", y = "Count")

p1
p2  
p3
p4
```

```{r wind-abundance-scatter}
# Scatter plot of wind vs abundance
ggplot(analysis_data, aes(x = mean_wind_speed, y = total_butterflies)) +
  geom_point(alpha = 0.5, aes(color = factor(view_id))) +
  scale_y_log10() +
  geom_smooth(method = "gam", se = TRUE, color = "black") +
  geom_vline(xintercept = 2, color = "red", linetype = "dashed", alpha = 0.7) +
  labs(title = "Raw Relationship: Wind Speed vs Butterfly Abundance",
       x = "Mean Wind Speed (m/s)", 
       y = "Total Butterflies (log scale)",
       color = "View ID") +
  theme(legend.position = "bottom")
```

# GAM Analysis

## Model 1: Basic GAM with Smooth Terms

```{r gam-model1}
# Fit GAM with smooth terms for wind variables
# Using negative binomial family for count data with overdispersion
gam1 <- gam(
  total_butterflies ~ 
    s(log_abundance_lag1, k = 10) +
    s(mean_wind_speed, k = 10) +
    s(temperature, k = 8) +
    s(sunlight_prop, k = 6) +
    s(hour_of_day, k = 8, bs = "cc") +  # Cyclic spline for hour of day
    factor(view_id),
  family = nb(),
  data = analysis_data,
  method = "REML"
)

summary(gam1)
```

## Model 2: GAM with Interaction Terms

```{r gam-model2}
# GAM with wind-temperature interaction
gam2 <- gam(
  total_butterflies ~ 
    s(log_abundance_lag1, k = 10) +
    s(mean_wind_speed, k = 10) +
    s(temperature, k = 8) +
    ti(mean_wind_speed, temperature, k = c(6, 6)) +  # Tensor interaction
    s(sunlight_prop, k = 6) +
    s(hour_of_day, k = 8, bs = "cc") +
    factor(view_id),
  family = nb(),
  data = analysis_data,
  method = "REML"
)

summary(gam2)
```

## Model 3: Alternative Wind Metrics

```{r gam-model3}
# Test different wind metrics
gam3 <- gam(
  total_butterflies ~ 
    s(log_abundance_lag1, k = 10) +
    s(max_wind_speed, k = 10) +  # Maximum wind instead of mean
    s(wind_variance, k = 8) +    # Wind variability (gustiness)
    s(temperature, k = 8) +
    s(sunlight_prop, k = 6) +
    s(hour_of_day, k = 8, bs = "cc") +
    factor(view_id),
  family = nb(),
  data = analysis_data,
  method = "REML"
)

summary(gam3)
```

# Model Comparison and Selection

```{r model-comparison}
# Compare models using AIC
model_comparison <- data.frame(
  Model = c("GAM1: Basic smooth terms", 
            "GAM2: Wind-temperature interaction", 
            "GAM3: Max wind + variance"),
  AIC = c(AIC(gam1), AIC(gam2), AIC(gam3)),
  Dev_Explained = c(
    paste0(round(summary(gam1)$dev.expl * 100, 1), "%"),
    paste0(round(summary(gam2)$dev.expl * 100, 1), "%"),
    paste0(round(summary(gam3)$dev.expl * 100, 1), "%")
  )
)

model_comparison %>%
  arrange(AIC) %>%
  kable(caption = "Model Comparison Results")

# Select best model
best_model <- list(gam1, gam2, gam3)[[which.min(c(AIC(gam1), AIC(gam2), AIC(gam3)))]]
best_model_name <- c("GAM1", "GAM2", "GAM3")[which.min(c(AIC(gam1), AIC(gam2), AIC(gam3)))]

print(paste("Best model:", best_model_name))
```

# Model Diagnostics

```{r diagnostics-gam}
# Comprehensive diagnostics for best model
gam.check(best_model)
```

```{r diagnostic-plots}
# Custom diagnostic plots
# Residuals vs fitted
p1 <- ggplot() +
  geom_point(aes(x = fitted(best_model), y = residuals(best_model)), alpha = 0.6) +
  geom_smooth(aes(x = fitted(best_model), y = residuals(best_model)), 
              method = "loess", se = TRUE, color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values", y = "Residuals")

# Q-Q plot of residuals
p2 <- ggplot() +
  geom_qq(aes(sample = residuals(best_model))) +
  geom_qq_line(aes(sample = residuals(best_model)), color = "red") +
  labs(title = "Q-Q Plot of Residuals",
       x = "Theoretical Quantiles", y = "Sample Quantiles")

# Residuals vs wind speed
p3 <- ggplot() +
  geom_point(aes(x = analysis_data$mean_wind_speed, y = residuals(best_model)), alpha = 0.6) +
  geom_smooth(aes(x = analysis_data$mean_wind_speed, y = residuals(best_model)), 
              method = "loess", se = TRUE, color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Wind Speed",
       x = "Mean Wind Speed (m/s)", y = "Residuals")

# Scale-location plot
p4 <- ggplot() +
  geom_point(aes(x = fitted(best_model), y = sqrt(abs(residuals(best_model)))), alpha = 0.6) +
  geom_smooth(aes(x = fitted(best_model), y = sqrt(abs(residuals(best_model)))), 
              method = "loess", se = TRUE, color = "red") +
  labs(title = "Scale-Location Plot",
       x = "Fitted Values", y = "√|Residuals|")

p1
p2
p3
p4
```

```{r temporal-diagnostics}
# Check for temporal autocorrelation in residuals
analysis_data_with_resids <- analysis_data %>%
  mutate(
    fitted_vals = fitted(best_model),
    residuals = residuals(best_model)
  )

# Plot residuals over time by deployment
ggplot(analysis_data_with_resids, aes(x = datetime, y = residuals)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~deployment_id, scales = "free_x") +
  labs(title = "Residuals Over Time by Deployment",
       x = "DateTime", y = "Residuals") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Effect Plots and Interpretation

```{r effect-plots}
# Plot smooth terms from best model
draw(best_model, pages = 1)
```

```{r individual-effects}
# Create individual effect plots
# Wind effect
wind_effect <- draw(best_model, select = "s(mean_wind_speed)") +
  geom_vline(xintercept = 2, color = "red", linetype = "dashed", alpha = 0.7) +
  labs(title = "Effect of Mean Wind Speed on Log Abundance",
       subtitle = "Red line shows 2 m/s threshold from Leong (2016)")

# Temperature effect  
temp_effect <- draw(best_model, select = "s(temperature)") +
  labs(title = "Effect of Temperature on Log Abundance")

# Lagged abundance effect
lag_effect <- draw(best_model, select = "s(log_abundance_lag1)") +
  labs(title = "Effect of Previous Abundance on Current Abundance")

# Hour of day effect
hour_effect <- draw(best_model, select = "s(hour_of_day)") +
  labs(title = "Daily Activity Pattern (Hour of Day Effect)")

wind_effect
temp_effect
lag_effect  
hour_effect
```

```{r predictions}
# Generate predictions to visualize wind effect
wind_pred_data <- expand_grid(
  mean_wind_speed = seq(0, max(analysis_data$mean_wind_speed), length.out = 100),
  log_abundance_lag1 = median(analysis_data$log_abundance_lag1),
  temperature = median(analysis_data$temperature),
  sunlight_prop = median(analysis_data$sunlight_prop),
  hour_of_day = 12,
  view_id = levels(factor(analysis_data$view_id))[1]
)

# Get predictions with confidence intervals
wind_predictions <- predict(best_model, wind_pred_data, se.fit = TRUE, type = "link")

wind_pred_plot <- wind_pred_data %>%
  mutate(
    predicted = exp(wind_predictions$fit),  # Back-transform from log scale
    lower = exp(wind_predictions$fit - 1.96 * wind_predictions$se.fit),
    upper = exp(wind_predictions$fit + 1.96 * wind_predictions$se.fit)
  )

ggplot(wind_pred_plot, aes(x = mean_wind_speed)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, fill = "blue") +
  geom_line(aes(y = predicted), color = "blue", size = 1.2) +
  geom_vline(xintercept = 2, color = "red", linetype = "dashed", alpha = 0.7) +
  labs(title = "Predicted Butterfly Abundance vs Wind Speed",
       subtitle = "Holding other variables at median values",
       x = "Mean Wind Speed (m/s)",
       y = "Predicted Butterfly Count",
       caption = "Red dashed line: 2 m/s threshold") +
  scale_y_log10()
```

# Results Summary and Next Steps

```{r results-summary}
# Extract key statistics
wind_smooth_pvalue <- summary(best_model)$s.table["s(mean_wind_speed)", "p-value"]
wind_edf <- summary(best_model)$s.table["s(mean_wind_speed)", "edf"]
model_dev_expl <- summary(best_model)$dev.expl

cat("## Key Results:\n")
cat(paste("- Best model:", best_model_name, "\n"))
cat(paste("- Model explains", round(model_dev_expl * 100, 1), "% of deviance\n"))
cat(paste("- Wind smooth term p-value:", format(wind_smooth_pvalue, scientific = TRUE), "\n"))
cat(paste("- Wind smooth effective degrees of freedom:", round(wind_edf, 2), "\n"))
cat(paste("- Number of observations:", nrow(analysis_data), "\n"))
cat(paste("- Number of deployments:", n_distinct(analysis_data$deployment_id), "\n"))
```

## Model Performance Metrics

```{r performance-metrics}
# Additional performance metrics
r2_val <- cor(fitted(best_model), analysis_data$total_butterflies)^2
rmse_val <- sqrt(mean((fitted(best_model) - analysis_data$total_butterflies)^2))

cat("\n## Performance Metrics:\n")
cat(paste("- R-squared (fitted vs observed):", round(r2_val, 3), "\n"))
cat(paste("- RMSE:", round(rmse_val, 1), "butterflies\n"))
cat(paste("- AIC:", round(AIC(best_model), 1), "\n"))
```

## Next Steps and Recommendations

Based on this exploratory GAM analysis, here are the recommended next steps:

1. **Model Assumptions**: The GAM diagnostic plots should be examined carefully to ensure model assumptions are met.

2. **Threshold Testing**: The smooth function for wind speed can reveal whether there's evidence for a threshold effect around 2 m/s.

3. **Alternative Distributions**: If residual patterns suggest issues, consider zero-inflated models or different distribution families.

4. **Temporal Structure**: Consider adding AR(1) correlation structure if temporal autocorrelation persists in residuals.

5. **Random Effects**: The current model uses fixed effects for view_id. Consider switching to random effects if you want to generalize beyond these specific sites.

6. **Cross-validation**: Implement leave-one-deployment-out cross-validation to test predictive performance.

7. **Effect Sizes**: Focus on biological significance of effects, not just statistical significance.

8. **Mechanistic Models**: Based on GAM results, consider more structured models that incorporate specific hypotheses about wind thresholds.