---
title: "H2 Analysis: Wind as a General Disruptive Force"
subtitle: "Testing Wind Effects Beyond Specific Thresholds"
author: "Kyle Nessen"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    fig-width: 8
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

```{r setup}
#| include: false
# Load required libraries
library(tidyverse)
library(glmmTMB)
library(DHARMa)
library(performance)
library(emmeans)
library(ggeffects)
library(patchwork)
library(knitr)
library(kableExtra)
library(here)

# Set theme for plots
theme_set(theme_minimal(base_size = 12) +
          theme(panel.grid.minor = element_blank(),
                strip.background = element_rect(fill = "grey95", color = NA)))

# Color palette
pal <- c("#2E86AB", "#A23B72", "#F18F01", "#C73E1D", "#6A994E")
```

# Executive Summary

This analysis tests **Hypothesis 2**: that wind acts as a general disruptive force on monarch butterfly abundance, independent of specific threshold values. Moving beyond the 2 m/s threshold tested in H1, we examine whether different characteristics of wind exposure (sustained winds, gusts, wind variability) affect monarch roost dynamics using negative binomial generalized linear mixed models.

**Key Finding**: We find **minimal evidence** that wind acts as a general disruptive force on monarch abundance. Across multiple wind metrics and model specifications, wind effects remain statistically non-significant and practically negligible. This supports the H1 finding that monarchs are more resilient to wind exposure than commonly assumed.

# Introduction

## Research Question

Does wind exposure, regardless of specific threshold values, act as a disruptive force that reduces monarch butterfly abundance at overwintering roost sites?

## H2 Hypothesis Framework

Following our hierarchical hypothesis structure:

1. **H1**: Wind speeds exceeding 2 m/s disrupt monarch clustering → **No support found**
2. **H2**: Wind acts as a disruptive force on monarch abundance (this analysis)
3. **H3**: Wind effects scale with intensity
4. **H4**: Wind magnitude influences roost abandonment probability  
5. **H5**: Disruptive wind events affect long-term site fidelity

## Rationale for H2

The lack of support for the specific 2 m/s threshold in H1 doesn't necessarily mean wind has no effect. H2 tests whether:

- Different wind characteristics (sustained vs. gusts vs. variability) matter
- Wind effects exist but are not captured by threshold approaches
- Continuous wind metrics reveal relationships missed by binary thresholds

## Methodological Improvements from H1

Based on lessons learned from H1 analysis:

1. **Response variable**: Use raw abundance counts with negative binomial GLMMs instead of problematic proportional change transformations
2. **Wind metrics**: Test multiple continuous wind measures to identify which matters most
3. **Random effects**: Use `view_id` and `labeler` as established in H1 final analysis
4. **Temporal autocorrelation**: Control with lagged abundance term
5. **Model comparison**: Compare models with same response variable using AIC

# Methods

## Data Preparation

```{r data-prep}
# Run data preparation script
source(here("hypotheses", "H2-analysis", "H2_H3_data_prep.R"))

# Load enhanced dataset
data_path <- here("results", "H2_H3_enhanced_data.rds")

if (!file.exists(data_path)) {
  stop("Enhanced data not found. Please run H2_H3_data_prep.R first.")
}

df <- read_rds(data_path)

# Display data structure
summary_stats <- df %>%
  summarise(
    n_obs = n(),
    n_views = n_distinct(view_id),
    n_labelers = n_distinct(labeler),
    mean_abundance_t = mean(abundance_index_t),
    sd_abundance_t = sd(abundance_index_t),
    prop_zeros = mean(abundance_index_t == 0),
    mean_sustained_wind = mean(sustained_minutes_above_2ms),
    mean_gust_wind = mean(gust_minutes_above_2ms),
    mean_wind_variability = mean(wind_variability)
  )

kable(summary_stats, 
      caption = "H2 Analysis Dataset Summary",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Statistical Approach

### Why Negative Binomial GLMMs for H2?

Our approach addresses the key issues identified in H1:

- **Count data**: Abundance is non-negative integers
- **Overdispersion**: Variance exceeds mean in aggregated organisms
- **Zero inflation**: Many observations with zero butterflies
- **Temporal autocorrelation**: Include lagged abundance as predictor
- **Site variation**: Random effects for `view_id` and `labeler`

### Model Structure

$$\log(\mu_t) = \beta_0 + \beta_1 \cdot \text{Wind\_Metric} + \beta_2 \cdot \text{Temp} + \beta_3 \cdot \text{Sun} + \beta_4 \cdot \log(\text{Count}_{t-1} + 1) + u_{\text{view}} + u_{\text{labeler}}$$

Where Wind_Metric represents different wind characteristics to be compared.

# Data Exploration

## Wind Metric Relationships

```{r wind-exploration, fig.height=10}
#| fig-cap: "Relationships between different wind metrics and abundance"

# Wind metric distributions
p1 <- ggplot(df, aes(x = sustained_minutes_above_2ms)) +
  geom_histogram(bins = 30, fill = pal[1], alpha = 0.7) +
  labs(x = "Sustained Wind Minutes > 2ms", y = "Count",
       title = "Distribution of Sustained Wind Exposure") +
  scale_y_continuous(expand = c(0, 0))

p2 <- ggplot(df, aes(x = gust_minutes_above_2ms)) +
  geom_histogram(bins = 30, fill = pal[2], alpha = 0.7) +
  labs(x = "Gust Wind Minutes > 2ms", y = "Count",
       title = "Distribution of Gust Wind Exposure") +
  scale_y_continuous(expand = c(0, 0))

p3 <- ggplot(df, aes(x = wind_variability)) +
  geom_histogram(bins = 30, fill = pal[3], alpha = 0.7) +
  labs(x = "Wind Variability (|Gust - Sustained|/30)", y = "Count",
       title = "Distribution of Wind Variability") +
  scale_y_continuous(expand = c(0, 0))

# Correlation between wind metrics
p4 <- ggplot(df, aes(x = sustained_minutes_above_2ms, y = gust_minutes_above_2ms)) +
  geom_point(alpha = 0.4, color = pal[4]) +
  geom_smooth(method = "lm", se = TRUE, color = pal[5]) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(x = "Sustained Wind Minutes", y = "Gust Wind Minutes",
       title = "Relationship: Sustained vs Gust Wind")

(p1 + p2) / (p3 + p4)
```

## Raw Relationships with Abundance

```{r abundance-relationships, fig.height=8}
#| fig-cap: "Raw relationships between wind metrics and butterfly abundance"

# Create wind exposure categories for visualization
df_viz <- df %>%
  mutate(
    sustained_cat = case_when(
      sustained_minutes_above_2ms == 0 ~ "No Wind",
      sustained_minutes_above_2ms <= 5 ~ "Low (1-5 min)",
      sustained_minutes_above_2ms <= 15 ~ "Medium (6-15 min)",
      TRUE ~ "High (>15 min)"
    ),
    sustained_cat = factor(sustained_cat, levels = c("No Wind", "Low (1-5 min)", "Medium (6-15 min)", "High (>15 min)"))
  )

p1 <- ggplot(df_viz, aes(x = sustained_cat, y = abundance_index_t)) +
  geom_boxplot(fill = pal[1], alpha = 0.7) +
  scale_y_log10() +
  labs(x = "Sustained Wind Category", y = "Abundance (log scale)",
       title = "Abundance by Sustained Wind Categories")

p2 <- ggplot(df, aes(x = sustained_minutes_above_2ms, y = abundance_index_t)) +
  geom_point(alpha = 0.3, color = pal[2]) +
  geom_smooth(method = "loess", se = TRUE, color = pal[3]) +
  scale_y_log10() +
  labs(x = "Sustained Wind Minutes", y = "Abundance (log scale)",
       title = "Abundance vs Continuous Sustained Wind")

p3 <- ggplot(df, aes(x = wind_variability, y = abundance_index_t)) +
  geom_point(alpha = 0.3, color = pal[4]) +
  geom_smooth(method = "loess", se = TRUE, color = pal[5]) +
  scale_y_log10() +
  labs(x = "Wind Variability", y = "Abundance (log scale)",
       title = "Abundance vs Wind Variability")

p4 <- ggplot(df, aes(x = log_lag_abundance, y = abundance_index_t)) +
  geom_point(alpha = 0.3, color = pal[3]) +
  geom_smooth(method = "loess", se = TRUE, color = pal[4]) +
  scale_y_log10() +
  labs(x = "Log Lagged Abundance", y = "Current Abundance (log scale)",
       title = "Temporal Autocorrelation")

(p1 + p2) / (p3 + p4)
```

# H2 Analysis: Model Comparison

## Testing Different Wind Characteristics

We fit separate models for each wind metric to avoid collinearity and identify which characteristic matters most.

```{r h2-models}
# Model 1: Sustained wind
m1_sustained <- glmmTMB(
  abundance_index_t ~ 
    log_lag_abundance + 
    sustained_wind_std +
    temp_std + 
    sun_std +
    (1 | view_id) + 
    (1 | labeler),
  data = df,
  family = nbinom2,
  control = glmmTMBControl(optimizer = nlminb)
)

# Model 2: Gust wind  
m2_gust <- glmmTMB(
  abundance_index_t ~ 
    log_lag_abundance + 
    gust_wind_std +
    temp_std + 
    sun_std +
    (1 | view_id) + 
    (1 | labeler),
  data = df,
  family = nbinom2,
  control = glmmTMBControl(optimizer = nlminb)
)

# Model 3: Wind variability
m3_variability <- glmmTMB(
  abundance_index_t ~ 
    log_lag_abundance + 
    wind_variability_std +
    temp_std + 
    sun_std +
    (1 | view_id) + 
    (1 | labeler),
  data = df,
  family = nbinom2,
  control = glmmTMBControl(optimizer = nlminb)
)

# Model 4: Multiple wind metrics (if correlations allow)
wind_correlations <- cor(df[c("sustained_wind_std", "gust_wind_std", "wind_variability_std")], 
                        use = "complete.obs")

# Only include multiple metrics if not highly correlated (r < 0.8)
if (max(abs(wind_correlations[wind_correlations != 1])) < 0.8) {
  m4_combined <- glmmTMB(
    abundance_index_t ~ 
      log_lag_abundance + 
      sustained_wind_std + 
      gust_wind_std +
      wind_variability_std +
      temp_std + 
      sun_std +
      (1 | view_id) + 
      (1 | labeler),
    data = df,
    family = nbinom2,
    control = glmmTMBControl(optimizer = nlminb)
  )
  include_combined <- TRUE
} else {
  message("Wind metrics highly correlated - skipping combined model")
  include_combined <- FALSE
}

# Model comparison
model_comparison <- data.frame(
  Model = c("Sustained Wind", "Gust Wind", "Wind Variability"),
  AIC = c(AIC(m1_sustained), AIC(m2_gust), AIC(m3_variability)),
  BIC = c(BIC(m1_sustained), BIC(m2_gust), BIC(m3_variability)),
  LogLik = c(logLik(m1_sustained), logLik(m2_gust), logLik(m3_variability))
)

if (include_combined) {
  model_comparison <- rbind(model_comparison,
                           data.frame(Model = "Combined",
                                     AIC = AIC(m4_combined),
                                     BIC = BIC(m4_combined), 
                                     LogLik = logLik(m4_combined)))
}

model_comparison <- model_comparison %>%
  arrange(AIC) %>%
  mutate(Delta_AIC = AIC - min(AIC))

kable(model_comparison, 
      caption = "H2: Model Comparison for Different Wind Characteristics",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Best Model Results

```{r best-model-results}
# Select best model (lowest AIC)
best_model_name <- model_comparison$Model[1]
best_model <- switch(best_model_name,
                    "Sustained Wind" = m1_sustained,
                    "Gust Wind" = m2_gust,
                    "Wind Variability" = m3_variability,
                    "Combined" = m4_combined)

# Extract and format coefficients
coef_table <- broom.mixed::tidy(best_model, conf.int = TRUE) %>%
  filter(effect == "fixed") %>%
  mutate(
    estimate_exp = exp(estimate),
    conf.low_exp = exp(conf.low),
    conf.high_exp = exp(conf.high),
    significant = p.value < 0.05
  )

# Format for presentation
coef_table %>%
  mutate(
    `Rate Ratio` = sprintf("%.3f", estimate_exp),
    `95% CI` = sprintf("[%.3f, %.3f]", conf.low_exp, conf.high_exp),
    `P-value` = sprintf("%.4f", p.value),
    Significant = ifelse(significant, "Yes", "No")
  ) %>%
  select(Term = term, `Rate Ratio`, `95% CI`, `P-value`, Significant) %>%
  kable(caption = paste("H2 Best Model Results:", best_model_name)) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Model summary
cat("Best performing wind metric:", best_model_name)
cat("\nModel AIC:", round(AIC(best_model), 2))
```

## Model Diagnostics

```{r h2-diagnostics, fig.height=12}
#| fig-cap: "Diagnostic plots for H2 best model"

# DHARMa residual diagnostics
sim_res <- simulateResiduals(best_model, n = 1000)

# Create diagnostic plots
par(mfrow = c(3, 2), mar = c(4, 4, 3, 1))

# Basic DHARMa plots
plot(sim_res, main = "Q-Q Plot")
plotResiduals(sim_res, form = fitted(best_model), main = "Residuals vs Fitted")

# Check residuals against predictors
wind_var <- switch(best_model_name,
                  "Sustained Wind" = df$sustained_wind_std,
                  "Gust Wind" = df$gust_wind_std, 
                  "Wind Variability" = df$wind_variability_std)

plotResiduals(sim_res, form = wind_var, 
              main = paste("Residuals vs", best_model_name))
plotResiduals(sim_res, form = df$temp_std, main = "Residuals vs Temperature")
plotResiduals(sim_res, form = df$sun_std, main = "Residuals vs Sunlight")

# Test dispersion
testDispersion(sim_res)
```

```{r temporal-diagnostics}
#| fig-cap: "Temporal autocorrelation check for H2 model"

# Check temporal autocorrelation in residuals
df$residuals <- residuals(best_model, type = "pearson")

# Plot ACF
acf_result <- acf(df$residuals, lag.max = 10, plot = FALSE)
plot(acf_result, main = "Autocorrelation Function of Residuals")

# Calculate correlation coefficient
cat("Residual autocorrelation at lag 1:", round(acf_result$acf[2], 3))
```

# Effect Visualization

## Predicted Effects

```{r h2-effects, fig.height=8}
#| fig-cap: "Predicted effects from H2 best model"

# Generate predictions for the best wind metric
wind_pred_var <- switch(best_model_name,
                       "Sustained Wind" = "sustained_wind_std",
                       "Gust Wind" = "gust_wind_std",
                       "Wind Variability" = "wind_variability_std")

wind_pred <- ggpredict(best_model, 
                       terms = paste0(wind_pred_var, " [-2:2]"),
                       condition = c(log_lag_abundance = mean(df$log_lag_abundance),
                                    temp_std = 0,
                                    sun_std = 0))

# Temperature predictions
temp_pred <- ggpredict(best_model,
                       terms = "temp_std [-2:2]",
                       condition = c(log_lag_abundance = mean(df$log_lag_abundance),
                                    sun_std = 0))

# Convert back to original scales for interpretation
temp_mean <- mean(df$ambient_temp, na.rm = TRUE)
temp_sd <- sd(df$ambient_temp, na.rm = TRUE)
temp_pred$x_original <- temp_pred$x * temp_sd + temp_mean

# Wind effect plot
p_wind <- ggplot(wind_pred, aes(x = x, y = predicted)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              fill = pal[1], alpha = 0.3) +
  geom_line(color = pal[1], size = 1.5) +
  geom_hline(yintercept = mean(df$abundance_index_t),
             linetype = "dashed", alpha = 0.5) +
  labs(x = paste("Standardized", str_replace_all(best_model_name, "_", " ")),
       y = "Predicted Abundance",
       title = paste("Effect of", best_model_name, "on Abundance"),
       subtitle = "Based on H2 analysis")

# Temperature effect plot  
p_temp <- ggplot(temp_pred, aes(x = x_original, y = predicted)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              fill = pal[2], alpha = 0.3) +
  geom_line(color = pal[2], size = 1.5) +
  geom_hline(yintercept = mean(df$abundance_index_t),
             linetype = "dashed", alpha = 0.5) +
  labs(x = "Temperature (°C)",
       y = "Predicted Abundance", 
       title = "Effect of Temperature on Abundance",
       subtitle = "Consistent strong positive effect")

p_wind / p_temp
```

## Model Performance

```{r h2-performance}
#| fig-cap: "H2 model fit assessment"

# Get predictions and calculate pseudo R-squared
df$predicted <- predict(best_model, type = "response")

# Calculate pseudo R-squared
ss_res <- sum((df$abundance_index_t - df$predicted)^2)
ss_tot <- sum((df$abundance_index_t - mean(df$abundance_index_t))^2)
r2_pseudo <- 1 - (ss_res / ss_tot)

# Observed vs predicted plot
ggplot(df, aes(x = predicted, y = abundance_index_t)) +
  geom_point(alpha = 0.3, color = pal[3]) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", se = TRUE, color = pal[4]) +
  labs(x = "Predicted Abundance",
       y = "Observed Abundance",
       title = "H2 Model Fit: Observed vs. Predicted",
       subtitle = sprintf("Pseudo R² = %.3f", r2_pseudo)) +
  coord_fixed() +
  theme_minimal(base_size = 12)

# Performance summary
cat("Model Performance Summary:\n")
cat("Pseudo R-squared:", round(r2_pseudo, 3), "\n")
cat("RMSE:", round(sqrt(mean((df$abundance_index_t - df$predicted)^2)), 2), "\n")
cat("Mean Absolute Error:", round(mean(abs(df$abundance_index_t - df$predicted)), 2), "\n")
```

# Sensitivity Analysis

## Alternative Specifications

```{r h2-sensitivity}
# Test binary wind effect
m_binary <- glmmTMB(
  abundance_index_t ~ 
    log_lag_abundance + 
    any_sustained_wind +
    temp_std + 
    sun_std +
    (1 | view_id) + 
    (1 | labeler),
  data = df,
  family = nbinom2,
  control = glmmTMBControl(optimizer = nlminb)
)

# Test wind-temperature interaction
m_interaction <- glmmTMB(
  abundance_index_t ~ 
    log_lag_abundance + 
    sustained_wind_std * temp_std +
    sun_std +
    (1 | view_id) + 
    (1 | labeler),
  data = df,
  family = nbinom2,
  control = glmmTMBControl(optimizer = nlminb)
)

# Compare sensitivity models
sensitivity_comparison <- data.frame(
  Model = c("Best H2 Model", "Binary Wind", "Wind×Temperature"),
  AIC = c(AIC(best_model), AIC(m_binary), AIC(m_interaction)),
  Wind_Effect_P = c(
    coef(summary(best_model))$cond[2, "Pr(>|z|)"],
    coef(summary(m_binary))$cond["any_sustained_windTRUE", "Pr(>|z|)"],
    coef(summary(m_interaction))$cond["sustained_wind_std", "Pr(>|z|)"]
  )
) %>%
  arrange(AIC) %>%
  mutate(Delta_AIC = AIC - min(AIC))

kable(sensitivity_comparison, 
      caption = "H2 Sensitivity Analysis: Alternative Model Specifications",
      digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Discussion

## H2 Summary of Findings

Based on comprehensive testing of different wind characteristics:

```{r h2-summary}
# Extract key statistics from best model
wind_coef_row <- which(grepl("wind", coef_table$term))
if (length(wind_coef_row) > 0) {
  wind_effect <- coef_table$estimate[wind_coef_row]
  wind_rr <- coef_table$estimate_exp[wind_coef_row] 
  wind_p <- coef_table$p.value[wind_coef_row]
  wind_significant <- wind_p < 0.05
} else {
  wind_effect <- NA
  wind_rr <- NA
  wind_p <- NA
  wind_significant <- FALSE
}

cat("H2 Key Findings:\n")
cat("Best wind metric:", best_model_name, "\n")
cat("Wind effect coefficient:", round(wind_effect, 4), "\n")
cat("Rate ratio:", round(wind_rr, 3), "\n")
cat("P-value:", round(wind_p, 3), "\n")
cat("Statistically significant:", wind_significant, "\n")
```

### Evidence for H2: Wind as Disruptive Force

1. **Minimal wind effects**: The best-performing wind metric shows non-significant effects (p > 0.05)
2. **Small effect sizes**: Rate ratios close to 1.0 indicate practically negligible impacts  
3. **Consistent with H1**: Confirms the pattern that wind doesn't substantially affect abundance
4. **Model comparison**: No wind metric dramatically outperforms others

### Temperature Remains Dominant

Consistent with H1 findings, temperature continues to show:
- Highly significant positive effects (p < 0.001)
- Larger effect sizes than any wind metric
- Consistent direction across all model specifications

## Biological Interpretation

The lack of support for H2 suggests:

1. **Monarch resilience**: Butterfly clusters may be more robust to wind exposure than hypothesized
2. **Microhabitat selection**: Monarchs likely choose well-protected roosting locations
3. **Behavioral adaptation**: Individual butterflies may adjust position within clusters during wind
4. **Scale mismatch**: Wind effects may operate at different temporal or spatial scales

## Limitations

1. **Wind measurement**: Single-point measurements may not capture spatial wind variation
2. **Temporal resolution**: 30-minute intervals may miss immediate or longer-term responses
3. **Missing wind characteristics**: Wind direction, turbulence patterns not assessed
4. **Sample constraints**: Limited to two sites and one season

# Conclusions

## H2 Hypothesis Assessment

**H2: Wind acts as a disruptive force on monarch abundance**

**Conclusion: Not Supported**

- No wind metric shows statistically significant effects on abundance
- Effect sizes are practically negligible across all specifications
- Results robust to different model formulations and wind measures
- Consistent with H1 threshold analysis findings

## Implications for Monarch Ecology

1. **Revised understanding**: Monarch overwintering behavior appears less wind-sensitive than assumed
2. **Conservation priorities**: Focus should shift toward thermal habitat requirements
3. **Climate resilience**: Populations may be more robust to changing wind patterns
4. **Research direction**: Investigation of other environmental factors warranted

## Connection to H3

This analysis sets up H3 (intensity scaling) by establishing that:
- General wind exposure shows minimal effects
- If scaling relationships exist, they involve subtle patterns
- Temperature consistently dominates environmental effects
- Count-based modeling approach is appropriate

The lack of clear linear wind effects in H2 motivates examining whether wind effects might be non-linear or threshold-dependent in H3.

# Appendix

## Full Model Output

```{r appendix}
# Complete model summary for best H2 model
summary(best_model)

# Variance components
VarCorr(best_model)

# Performance metrics
performance::model_performance(best_model)
```

## Session Information

```{r session-info}
sessionInfo()
```