---
title: "Monarch Abundance Analysis with GAM"
author: "Kyle Nessen"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## Introduction

This analysis examines changes in monarch abundance using a robust GAM (Generalized Additive Model) approach. We'll explore temporal patterns and potential drivers of monarch population dynamics.

## Setup

```{r setup}
library(tidyverse)
library(mgcv)
library(lubridate)
library(plotly)
library(knitr)
library(DT)
library(here)
```

## Data Loading

```{r load-data}
# Load the monarch analysis data
monarch_data <- read_csv(here("data","monarch_analysis_lag30min.csv"))

glimpse(monarch_data)
```

## Data Preparation

```{r data-prep}
# Check grove distribution
table(monarch_data$grove)

# Remove any rows with missing values in key variables and exclude SC_8
model_data <- monarch_data %>%
  filter(deployment_id != "SC_8",
         !is.na(butterfly_difference_cbrt),
         !is.na(total_butterflies_t_lag),
         !is.na(max_gust),
         !is.na(temperature_avg),
         !is.na(butterflies_direct_sun_t_lag),
         !is.na(deployment_id),
         !is.na(deployment_day),
         !is.na(Observer),
         !is.na(observation_order_within_day_t))

cat("Clean dataset has", nrow(model_data), "observations\n")
```

## Model Building and Selection

```{r model-setup}
library(nlme)

# Define the random effects structure and correlation
random_structure <- list(deployment_id = ~ 1, Observer = ~ 1, deployment_day = ~ 1)
correlation_structure <- corAR1(form = ~ observation_order_within_day_t | deployment_day)

# Model specifications for AIC comparison
model_specs <- list(
  # Null model
  "M0_null" = "butterfly_difference_cbrt ~ total_butterflies_t_lag",
  
  # Single variable models
  "M1_gust" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + max_gust",
  "M2_temp" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + temperature_avg", 
  "M3_sun" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + butterflies_direct_sun_t_lag",
  
  # Two-variable combinations
  "M4_gust_temp" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + max_gust + temperature_avg",
  "M5_gust_sun" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + max_gust + butterflies_direct_sun_t_lag",
  "M6_temp_sun" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + temperature_avg + butterflies_direct_sun_t_lag",
  
  # Three-variable model (main effects only)
  "M7_all_main" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + max_gust + temperature_avg + butterflies_direct_sun_t_lag",
  
  # Two-way interactions
  "M8_gust_temp_int" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + max_gust * temperature_avg",
  "M9_gust_sun_int" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + max_gust * butterflies_direct_sun_t_lag",
  "M10_temp_sun_int" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + temperature_avg * butterflies_direct_sun_t_lag",
  
  # Two-way interactions with third variable as main effect
  "M11_gust_temp_int_plus_sun" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + max_gust * temperature_avg + butterflies_direct_sun_t_lag",
  "M12_gust_sun_int_plus_temp" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + max_gust * butterflies_direct_sun_t_lag + temperature_avg",
  "M13_temp_sun_int_plus_gust" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + temperature_avg * butterflies_direct_sun_t_lag + max_gust",
  
  # All two-way interactions
  "M14_all_two_way" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + max_gust * temperature_avg + max_gust * butterflies_direct_sun_t_lag + temperature_avg * butterflies_direct_sun_t_lag",
  
  # Three-way interaction
  "M15_three_way" = "butterfly_difference_cbrt ~ total_butterflies_t_lag + max_gust * temperature_avg * butterflies_direct_sun_t_lag",
  
  # Smooth terms models
  "M16_smooth_temp" = "butterfly_difference_cbrt ~ s(total_butterflies_t_lag) + s(temperature_avg) + butterflies_direct_sun_t_lag",
  "M17_smooth_sun" = "butterfly_difference_cbrt ~ s(total_butterflies_t_lag) + temperature_avg + s(butterflies_direct_sun_t_lag)",
  "M18_smooth_gust" = "butterfly_difference_cbrt ~ s(total_butterflies_t_lag) + s(max_gust) + temperature_avg + butterflies_direct_sun_t_lag",
  "M19_smooth_temp_sun" = "butterfly_difference_cbrt ~ s(total_butterflies_t_lag) + s(temperature_avg) + s(butterflies_direct_sun_t_lag)",
  "M20_smooth_all_main" = "butterfly_difference_cbrt ~ s(total_butterflies_t_lag) + s(max_gust) + s(temperature_avg) + s(butterflies_direct_sun_t_lag)",
  "M21_time_of_day" = "butterfly_difference_cbrt ~ s(total_butterflies_t_lag) + temperature_avg + butterflies_direct_sun_t_lag + s(time_within_day_t)",
  "M22_temp_time" = "butterfly_difference_cbrt ~ s(total_butterflies_t_lag) + s(temperature_avg) + butterflies_direct_sun_t_lag + s(time_within_day_t)",
  "M23_all_smooth_time" = "butterfly_difference_cbrt ~ s(total_butterflies_t_lag) + s(max_gust) + s(temperature_avg) + s(butterflies_direct_sun_t_lag) + s(time_within_day_t)"
)

cat("Total models to fit:", length(model_specs), "\n")
```

## Model Fitting

```{r model-fitting}
# Function to safely fit models
fit_model_safely <- function(formula_str, data) {
  tryCatch({
    formula_obj <- as.formula(formula_str)
    gamm(formula_obj, 
         data = data,
         random = random_structure,
         correlation = correlation_structure,
         method = "REML")
  }, error = function(e) {
    message("Failed to fit model: ", formula_str)
    message("Error: ", e$message)
    return(NULL)
  })
}

# Fit all models
cat("Fitting models...\n")
fitted_models <- map(model_specs, ~fit_model_safely(.x, model_data))

# Remove failed models
successful_models <- fitted_models[!map_lgl(fitted_models, is.null)]
cat("Successfully fitted", length(successful_models), "out of", length(model_specs), "models\n")
```

## Model Comparison

```{r model-comparison}
# Extract AIC values
aic_results <- map_dfr(names(successful_models), function(model_name) {
  model <- successful_models[[model_name]]
  data.frame(
    Model = model_name,
    Formula = model_specs[[model_name]],
    AIC = AIC(model$lme),
    LogLik = logLik(model$lme)[1],
    df = attr(logLik(model$lme), "df")
  )
}) %>%
  arrange(AIC) %>%
  mutate(
    Delta_AIC = AIC - min(AIC),
    AIC_weight = exp(-0.5 * Delta_AIC) / sum(exp(-0.5 * Delta_AIC))
  )

# Display results
aic_results %>%
  select(Model, AIC, Delta_AIC, AIC_weight, df) %>%
  kable(digits = 3, caption = "Model comparison by AIC")

# Show top 5 models
cat("\nTop 5 models by AIC:\n")
head(aic_results, 5) %>%
  select(Model, Formula, AIC, Delta_AIC) %>%
  kable(digits = 3)
```

## Best Model Analysis

```{r best-model-analysis}
# Get the best model
best_model_name <- aic_results$Model[1]
best_model <- successful_models[[best_model_name]]

cat("Best model:", best_model_name, "\n")
cat("Formula:", aic_results$Formula[1], "\n\n")

# Model summary
summary(best_model$gam)
```

## Effect Plots

```{r effect-plots}
# Plot smooth effects
par(mfrow = c(2, 2))
plot(best_model$gam, pages = 1, residuals = TRUE, pch = 19, cex = 0.3)

# Individual effect plots with better formatting
par(mfrow = c(2, 2))
plot(best_model$gam, select = 1, main = "Effect of Previous Butterfly Count", 
     xlab = "Total Butterflies (t-lag)", ylab = "Partial Effect", 
     residuals = TRUE, pch = 19, cex = 0.5)

plot(best_model$gam, select = 2, main = "Effect of Temperature", 
     xlab = "Average Temperature (°C)", ylab = "Partial Effect", 
     residuals = TRUE, pch = 19, cex = 0.5)

plot(best_model$gam, select = 3, main = "Diurnal Pattern", 
     xlab = "Time Within Day (minutes)", ylab = "Partial Effect", 
     residuals = TRUE, pch = 19, cex = 0.5)

# Linear effect of sun exposure (parametric term)
sun_effect <- summary(best_model$gam)$p.table["butterflies_direct_sun_t_lag", ]
plot(model_data$butterflies_direct_sun_t_lag, 
     model_data$butterflies_direct_sun_t_lag * sun_effect["Estimate"],
     xlab = "Butterflies in Direct Sun (t-lag)", 
     ylab = "Linear Effect",
     main = "Effect of Sun Exposure (Linear)",
     pch = 19, cex = 0.5, col = "blue")
abline(h = 0, lty = 2, col = "red")
```

```{r effect-interpretation}
# Create nicer ggplot versions
library(gratia)

# All smooth effects in one plot
draw(best_model$gam) + 
  theme_minimal() +
  labs(title = "Smooth Effects in Best Model")
```

```{r model-diagnostics}
# Model diagnostics
par(mfrow = c(2, 2))

# Residual plots using the lme component
plot(best_model$lme)

# Additional residual checks
residuals_df <- data.frame(
  fitted = fitted(best_model$lme),
  residuals = residuals(best_model$lme, type = "normalized")
)

# Q-Q plot
qqnorm(residuals_df$residuals, main = "Normal Q-Q Plot of Residuals")
qqline(residuals_df$residuals)

# Histogram of residuals
hist(residuals_df$residuals, main = "Distribution of Residuals", xlab = "Residuals")
```

```{r coefficient-interpretation}
# Extract and display coefficients
coef_table <- summary(best_model$gam)$p.table
coef_df <- data.frame(
  Variable = rownames(coef_table),
  Estimate = coef_table[, "Estimate"],
  SE = coef_table[, "Std. Error"],
  t_value = coef_table[, "t value"],
  p_value = coef_table[, "Pr(>|t|)"]
) %>%
  mutate(
    Significant = ifelse(p_value < 0.001, "***",
                        ifelse(p_value < 0.01, "**", 
                              ifelse(p_value < 0.05, "*", 
                                    ifelse(p_value < 0.1, ".", ""))))
  )

coef_df %>%
  kable(digits = 4, caption = "Coefficient estimates for best model")
```

```{r model-performance}
# Model performance metrics
cat("Model Performance:\n")
cat("AIC:", AIC(best_model$lme), "\n")
cat("Log-likelihood:", logLik(best_model$lme)[1], "\n")
cat("R-squared (approximate):", summary(best_model$gam)$r.sq, "\n")
cat("Deviance explained:", summary(best_model$gam)$dev.expl * 100, "%\n")

# Random effects variance
cat("\nRandom Effects Variance Components:\n")
print(VarCorr(best_model$lme))
```

## Results Summary

The analysis tested 15 different model configurations to determine which environmental factors best predict changes in monarch butterfly abundance. The models included various combinations of wind speed (max_gust), temperature (temperature_avg), and solar exposure (butterflies_direct_sun_t_lag), along with their interactions.

Key findings will be interpreted based on the best-performing model according to AIC criteria.

